\renewcommand{\abstractname}{Acknowledgements}
\begin{abstract}
    It's a common aphorism to state that research outcomes ``stand on the shoulders of giants'',
    referring to all the research and hard work that the outcomes build upon. It's definitely true,
    however I'd like to personally thank those who are rarely recognised: the carers, the cleaners, the
    cooks, the friends and family, the physiotherapists, the detectives, therapists, and all of the
    people who form a support group around a person. It takes a village to raise a child, and the
    same goes for a PhD. In both cases, there's a lot of thankless, gruelling work, which might
    sound similar, but you don't get a certificate and floppy hat at the end. You're all giants to
    me. 
     
    I'd like to thank the following people for their help and support: my partner Kara Hatherly,
    Debbie \& Geoff Fear, Biara Webster, Ryan Williams and his team, Patricia Durning and Gary
    Penhall. I would \emph{never} have finished this project without you. 
     
    I thank my supervisor, Gernot Heiser, for encouraging me, giving me the time I needed, and
    ultimately proving the support that led to me complete the project in the face of extreme
    external events. My co-supervisor, Kevin
    Elphinstone, for being drawn into hours of whiteboard discussion, after I waited earnestly at
    the coffee machine.  Additionally all of the students, engineers and researchers who have helped
    at some point or another: David Greenaway, Thomas Sewell, Adrian Danis, Kent McLeod, Yanyan
    Shen, Luke Mondy, Hesham Almatary and Qian Ge. Further thanks to everyone in the Trustworthy Systems
    team, past and present, you're all fantastic.
    
    Matthew Fernandez, Rafal Kolanski, and Joel Beeren get a special thanks not only for help with
    the work, but for supporting me in the lab, and the pub, and being my friends even at my worst,
    you were a huge factor in keeping me sane. 
    
    Finally I'd like to thank my veritable army of proof readers: Kara Hatherly, Kent McLeod, 
    Peter Chubb,  Tim Cerexhe, Alex Legg, Joel Beeren, Sebastion Holzapfel, Robert Sison,  
    Sean Peters, Ihor Kuz and of course, Gernot Heiser.
\end{abstract}
\addcontentsline{toc}{chapter}{Acknowledgements}
\clearpage
